Judicious Batching: The Impact of Batch Ordering Imaging Tests in the Emergency Department

### 2024-09-09 manuscript/manuscript v2.Rmd
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variabilityin the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits, we find that quasi-random assignment to a top (versus bottom) decile batching physician significantly increases excess diagnostic testing. Instrumental variable results show that batching leads to an additional 13.5 tests per 100 batches, resulting in excess tests that would not have been ordered under a sequential-ordering strategy. Contrary to expectations, batch ordering does not reduce ED length of stay or affect 72-hour return rates, challenging the perceived efficiency of batching and highlighting the necessity for more targeted testing strategies in emergency care.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
---

@@

With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly. We find that being treated by a top decile batching physician, compared to a bottom decile batching physician, results in an additional 13.5 tests per 100 patient encounters. We find no evidence that batch-ordering reduces ED length of stay or 72-hour return rates. These results suggest that the benefits of batch-ordering are limited, and that the practice may be associated with significant costs. Our findings are robust to analyses probing the key assumptions of our batching definition.

Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

@@

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe approximately $48,000$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $10$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include urinary complaints and allergic reactions. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving physicians who treat over 500 ED cases per year.

@@

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. Below, we detail our primary outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.

@@

  *(c) 72-hour return*-- The rate of patients returning to the ED within 72 hours of their initial visit is a proxy for the quality of care received. A high 72-hour return rate might indicate inadequate treatment or diagnostic oversight during the initial visit. We measure this outcome as the proportion of patients with a subsequent ED visit within 72 hours post-initial discharge. The hypothesis is that batch testing may lead to variations in the 72-hour return rate, potentially influencing the quality of care received.

@@

We define ``batching" in line with standard emergency medicine practices. It involves a physician ordering a comprehensive set of diagnostic tests at once, typically covering a broad range of potential diagnoses. This contrasts with non-batching, where tests are ordered more selectively based on the information available at the time, with additional tests potentially ordered later as needed.

We operationalize batching as the ordering of multiple diagnostic tests within a 5-minute window. Sensitivity analyses on this cutoff point showed that our results are robust to this definition. This operational definition aligns with the concept of batching as a single comprehensive effort. For our analysis, all lab-based testing ordered in this timeframe are categorized as one distinct type of testing (lab). In contrast, each imaging test (e.g., X-ray, CT scan) are considered a separate distinct test. Therefore, a batch in our study is defined as consisting of two or more distinct diagnostic tests, which could be a combination of lab and imaging tests or multiple imaging tests.

This approach reflects the realistic diagnostic strategies in an ED setting and allows for a clear distinction between batching and non-batching behaviors. The differentiation between lab tests as one collective unit and imaging tests as individual units is based on their operational impact in the ED, particularly regarding patient waiting time and processing queues.

@@

\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview of the emergency department (ED) characteristics, patient demographics, and medical tests for our baseline sample, derived from the data collected during the study period. On average, the emergency department manages a volume of approximately 24 patients (Mean = 24.18), indicating a significant but manageable patient load. Key physiological markers such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are prevalent among patients, albeit at varying degrees, highlighting critical aspects of emergency care. Notably, a small proportion of patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01) conditions, emphasizing the diversity of cases encountered in the ED setting. The Emergency Severity Index (ESI) averages at 2.78, suggesting that most cases fall within the moderate acuity level.

Common chief complaints in the emergency department paint a vivid picture of patient needs and healthcare demands. The data reveals abdominal complaints (Mean = 0.14), extremity complaints (Mean = 0.12), and chest pain (Mean = 0.08) as the most frequently occurring, followed by neurological (Mean = 0.08) and gastrointestinal issues (Mean = 0.08). This information is crucial in understanding the primary reasons for ED visits and the required resources for patient care.

The patient characteristics in our sample depict a diverse demographic landscape. The average arrival age in the ED is approximately 58 years (Mean = 58.33), indicating a predominantly middle-aged population. The racial composition is predominantly White (Mean = 0.89), with smaller proportions of Black (Mean = 0.04) and Asian (Mean = 0.03) patients. Notably, a slight majority of the patients are female (Mean = 0.54), offering insights into gender dynamics in healthcare utilization.

In terms of diagnostic tests, the data reveals a high reliance on lab tests (Mean = 0.77), indicating their crucial role in patient diagnosis and management. X-Ray (Mean = 0.47) and CT scans, both non-contrast (Mean = 0.21) and contrast (Mean = 0.19), are also frequently employed, underscoring the importance of imaging in modern emergency medicine. Ultrasound usage (Mean = 0.11) and the practice of batch ordering tests (Mean = 0.38) further reflect the operational aspects of the ED and the strategies employed to manage patient care effectively.

\begin{table}[ht]
\centering
\caption{Summary Statistics}
\label{tab:summary_statistics}
\begin{tabular}{p{10.5cm}cccc}
\toprule
\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
\midrule
\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
ED Volume & 24.18 & 15 & 25 & 32 \\
Tachycardic & 0.19 & & & \\
Tachypneic & 0.09 & & & \\
Febrile & 0.02 & & & \\
Hypotensive & 0.01 & & & \\ 
ESI & 2.78 & 2 & 3 & 3 \\
Complaint: Abdominal & 0.14 & & & \\
Complaint: Extremity & 0.12 & & & \\
Complaint: Chest Pain & 0.08 & & & \\
Complaint: Neurological & 0.08 & & & \\
Complaint: Gastrointestinal & 0.08 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
Arrival Age & 58.33 & 44 & 62 & 75 \\
Race: White & 0.89 & & & \\
Race: Black & 0.04 & & & \\
Race: Asian & 0.03 & & & \\
Gender: Female & 0.54 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Tests}} \\
X-Ray & 0.47 & & & \\
Ultrasound & 0.11 & & & \\
Non-Contrast CT & 0.21 & & & \\
Contrast CT & 0.19 & & & \\
Lab & 0.77 & & & \\
Tests were Batch Ordered & 0.38 & & & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \caption{Balancing Test: Wald Test for Equality of Means}
    \label{tab:wald_test}
    \begin{tabular}{p{10cm}cccc}
        \toprule
        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 6232 & 2.587 & 0.108 \\ 
        Back or Flank Pain & 2552 & 1.637 & 0.201 \\ 
        Chest Pain & 3525 & 0.407 & 0.524 \\ 
        Extremity Complaints & 5265 & 1.847 & 0.174 \\ 
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2381 & 0.023 & 0.880 \\ 
        Gastrointestinal Issues & 3323 & 0.105 & 0.746 \\ 
        Neurological Issue & 3495 & 0.135 & 0.713 \\ 
        Shortness of Breath & 2966 & 1.324 & 0.250 \\ 
        Skin Complaints & 2178 & 0.383 & 0.536 \\ 
        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
        \midrule
        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 13914 & 0.011 & 0.915 \\ 
        ESI 3, 4, or 5 & 29386 & 0.010 & 0.921 \\ 
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{table}

@@

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint} + \varepsilon_{i,t}
\end{equation}

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on date $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision. As stated earlier, these controls are what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure \ref{fig:physician_batching} graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically. \hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $8$pp increase in the likelihood of having tests batch-ordered in the ED. The F-statistic is 50 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.

\begin{figure}[h]
  \centering
  \caption{Batch Frequency by Physician for High Volume Complaints}
  \label{fig:physician_batching}
  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
\begin{tablenotes}
\tiny
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests. Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue). The quartile of batchers is calculated based on the batching rate across all complaint areas per physician. Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching. Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
\end{tablenotes}  
\end{figure}

\begin{table}[ht]
\centering
\caption{First Stage: Effect of Batch Tendency on Test Batching}
\label{tab:regression}
\begin{tabular}{p{9cm}ccc}
\toprule
\textbf{} & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 1.06 *** & 1.08 *** & 1.08 *** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43327 & 43327 & 43327 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\tiny
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

This reduced form will allow us to check that our instrument is a strong instrument. To study the effects of test batching in the ED on an outcome $Y_i$, we estimate the following 2SLS equations using our baseline sample:

@@

\begin{equation}
Batched_i = \delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i + \nu_i
\end{equation}

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization. 
and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

@@

In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department. Our reduced form regression results, as presented in Model 1 of Table \ref{tab:reducedform}, reveal that batch tendency correlates with prolonged ED length of stay (LOS), increased test ordering, and does not materially alter the likelihood of a 72-hour return. Notably, these associations are nuanced by an underlying factor: the physician's testing inclination.

>>>>

---
title: "Judicious Batching: The Impact of Batch Ordering Imaging Tests in the Emergency Department"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits, we find that quasi-random assignment to a top (versus bottom) decile batching physician significantly increases excess diagnostic testing. Instrumental variable results show that batching leads to an additional 13.5 tests per 100 batches, resulting in excess tests that would not have been ordered under a sequential-ordering strategy. Contrary to expectations, batch ordering does not reduce ED length of stay or affect 72-hour return rates, challenging the perceived efficiency of batching and highlighting the necessity for more targeted testing strategies in emergency care.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---

@@

With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly. Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

@@

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe approximately $48,000$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $2$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $36,765$ encounters involving full-time physicians who treat over 500 ED cases per year.

@@

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. Below, we detail our primary outcomes: (a) ED length of stay (LOS), and (b) resource utilization.

@@

*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This outcome is used to assess the quality of care provided during the initial ED visit. The hypothesis is that batch testing may lead to variations in the quality of care provided, potentially influencing the likelihood of a return visit.

@@

We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window. In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study. Therefore, a batch in our study consists of two or more distinct imaging tests.

@@

hyperref[tab]{Table 1} provides a concise overview of emergency department (ED) characteristics, patient demographics, and medical tests based on our study data. The ED maintains an average volume of 24.3 patients, reflecting a significant yet manageable workload. Key physiological indicators include tachycardia (18.8%) and tachypnea (9.51%), showcasing the critical conditions often treated in emergency settings. Additionally, fever and hypotension are observed in 2.43% and 1.57% of patients, respectively, highlighting the variety of urgent care demands.

Patient complaints commonly involve abdominal issues (16.9%), extremity problems (14.3%), and chest pain (9.58%), along with neurological (9.50%) and gastrointestinal complaints (9.04%). These figures emphasize the frequent health concerns that drive ED visits and the resources needed for effective response.

The patient demographic profile shows an average age of 58.3 years, predominantly middle-aged and elderly individuals. Racial distribution is mainly White (88.6%), with smaller percentages of Black (4.12%) and Asian (2.97%) populations. A slight majority of patients are female (54.5%), indicating gender trends in healthcare utilization.

Diagnostic procedures heavily rely on lab tests (78.5%), essential for thorough patient evaluations. Imaging tests such as X-rays (48.1%) and CT scans—both non-contrast (23.1%) and contrast (20.5%)—play pivotal roles in diagnostics. The use of ultrasounds (12.5%) and the strategy of batch ordering tests (6.32%) are critical operational aspects of the ED, reflecting approaches to efficient patient management.

\begin{table}[ht]
\centering
\caption{Summary Statistics of Emergency Department Encounters}
\label{tab:summary_statistics}
\begin{tabular}{p{10.5cm}cccc}
\toprule
\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
\midrule
\multicolumn{5}{l}{\textbf{Emergency Department Characteristics}} \\
Patients in ED & 24.3 & 16 & 25 & 33 \\
\\
Tachycardic & 0.188 & & & \\
Tachypneic & 0.095 & & & \\
Febrile & 0.024 & & & \\
Hypotensive & 0.016 & & & \\
\\
ESI & 2.74 & 2 & 3 & 3 \\
\\
Complaint: Abdominal Complaints & 0.169 & & & \\
Complaint: Extremity Complaints & 0.143 & & & \\
Complaint: Chest Pain & 0.096 & & & \\
Complaint: Neurological Issue & 0.095 & & & \\
Complaint: Gastrointestinal Issues & 0.090 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
Arrival Age & 58.3 & 44 & 61 & 75 \\
\\
Race: White & 0.886 & & & \\
Race: Black & 0.041 & & & \\
Race: Asian & 0.030 & & & \\
\\
Gender: Female & 0.545 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Tests}} \\
X-Ray & 0.481 & & & \\
Ultrasound & 0.125 & & & \\
Non-Contrast CT & 0.231 & & & \\
Contrast CT & 0.205 & & & \\
Lab & 0.785 & & & \\
\\
Imaging Tests were Batch Ordered & 0.063 & & & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \caption{Balancing Test: Wald Test for Equality of Means}
    \label{tab:wald_test}
    \begin{tabular}{p{10cm}cccc}
        \toprule
        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 6232 & 2.587 & 0.108 \\ 
        Back or Flank Pain & 2552 & 1.637 & 0.201 \\ 
        Chest Pain & 3525 & 0.407 & 0.524 \\ 
        Extremity Complaints & 5265 & 1.847 & 0.174 \\ 
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2381 & 0.023 & 0.880 \\ 
        Gastrointestinal Issues & 3323 & 0.105 & 0.746 \\ 
        Neurological Issue & 3495 & 0.135 & 0.713 \\ 
        Shortness of Breath & 2966 & 1.324 & 0.250 \\ 
        Skin Complaints & 2178 & 0.383 & 0.536 \\ 
        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
        \midrule
        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 13914 & 0.011 & 0.915 \\ 
        ESI 3, 4, or 5 & 29386 & 0.010 & 0.921 \\ 
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\small
\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{table}

@@

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \alpha_{lab} + \varepsilon_{i,t}
\end{equation}

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision. Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case. As stated earlier, these controls are more than what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure \ref{fig:physician_batching} graphs the frequency with which physicians batch order across different chief complaints, highlighting that the variation in batching differs starkly. We note that there is high correlation between

\begin{figure}[h]
  \centering
  \caption{Variation in Physician Batching}
  \label{fig:physician_batching}
  \includegraphics[width=4.25in]{../outputs/figures/Figure 1.png}
\begin{tablenotes}
\small
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.}
\end{tablenotes}  
\end{figure}

\hyperref[table:first_stage]{Table 1} illustrates the "first stage" results from both linear and probit models, emphasizing the robust relevance of the instrument across different statistical frameworks. In both modeling approaches, a 10 percentage point increase in the batch-tendency of physicians consistently results in a significant increase in the likelihood of batch ordering tests in the emergency department—1.15 percentage points in the linear model and 93.24 percentage points in the probit model. This consistent significance across models underscores the strength and relevance of batch tendency as an instrumental variable. Given the robustness to different model specifications, we move forward with a probit model in our first stage to estimate the effect of batch ordering on patient outcomes.

\begin{table}[!htbp] \centering 
  \caption{Comparison of First Stage Estimates: Linear vs. Probit Models}
  \label{table:first_stage}
\begin{tabular}{@{\extracolsep{5pt}}lcccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{Linear Model} & \multicolumn{3}{c}{Probit Model} \\ 
\cline{2-4} \cline{5-7}
\\[-1.8ex] & (1) & (2) & (3) & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Batch Tendency & 1.25$^{***}$ & 1.24$^{***}$ & 1.15$^{***}$ & 8.657$^{***}$ & 9.135$^{***}$ & 8.515$^{***}$ \\ 
  & (0.04) & (0.03) & (0.02) & (0.659) & (0.709) & (0.789) \\ 
  & & & & & & \\
\textit{Day of Week-Time of Day FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\textit{Month of Year FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
\textit{Complaint/Severity FE} & & $\checkmark$ & $\checkmark$ & & $\checkmark$ & $\checkmark$ \\
\textit{Laboratory Tests Ordered} & & & $\checkmark$ & & & $\checkmark$ \\
\hline 
\hline \\[-1.8ex] 
F Statistic & 13.13 & 26.02 & 29.38 &  &  &  \\
Akaike Inf. Crit. & & & & 16,957.210 & 15,657.200 & 15,349.280 \\ 
\\
\textit{N} & 36,765 & 36,765 & 36,765 & 36,765 & 36,765 & 36,765 \\

\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\small
\item \textit{Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

@@

This reduced form will allow us to check the strength of our instrument. To study the effects of test batching in the ED on an outcome $Y_i$, we employ a two-stage estimation approach using our baseline sample:

@@

\begin{equation}
P(Batched_i = 1) = \Phi(\delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i)
\end{equation}

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization, and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

@@

In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department. We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination. Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.

To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency. Specifically, we calculate the physician's residualized leave-out average test rate, which serves as a proxy for their propensity to order tests. It is important to note the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination. This substantial correlation suggests that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies. Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.

Reduced-form regression results in Table \ref{tab:reducedform} reveal the effect the average total effect of batch tendency on resource utilization and LOS. However, it is important to note that the estimated total effect of batch tendency accounts for both direct and indirect (mediated) effects of batch tendency. We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered. The increased testing volume, in turn, is associated with an extended LOS.

\begin{table}[!htbp] \centering 
\caption{Reduced-Form Results: Length of Stay and Number of Tests}
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & ln(LOS) & Number of Tests \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
 Batch Tendency & $-$0.109 & 0.149$^{***}$ \\ 
  & (0.464) & (0.048) \\ 
  & & \\ 
 Testing Inclination & 0.605$^{**}$ & 0.952$^{***}$ \\ 
  & (0.240) & (0.023) \\ 
  & & \\ 
Seasonality and shift fixed effects? & Yes & Yes \\
Chief Complaint? & Yes & Yes  \\
\hline 
\textit{N} & 41,929 & 41,929 \\ 
R$^{2}$ & 0.215 & 0.265 \\ 
Adjusted R$^{2}$ & 0.213 & 0.263 \\ 
Residual Std. Error (df = 41800) & 0.490 & 0.842 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\small
\item \textit{Estimates of the reduced form are for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

We introduce a Directed Acyclic Graph (DAG) to illustrate the interconnected pathways between batch tendency, batch ordering, and primary outcomes. The DAG in Figure \ref{fig:DAG} shows that batch tendency directly influences batch ordering, which in turn affects the number of tests ordered. The number of tests ordered, in turn, influences the length of stay. The DAG also illustrates the effect of batch tendency on the length of stay  is mediated by the number of tests ordered.

\begin{figure}[h]
  \centering
  \caption{Directed Acyclic Graph}
  \label{fig:DAG}
  \includegraphics[width=6in]{../outputs/figures/dagitty.png}
\begin{tablenotes}
\small
\item \textit{}
\end{tablenotes}  
\end{figure}
